Scalar Feedback based Joint Time-Frequency Precoder interpolation for MIMO-OFDM Systems

### 2019-10-15 paper/main.tex
Further changes by Kumar

In this paper, we propose a method for exploiting the joint time-frequency correlation of the precoder's scalar parameters on the transmitter using CSI feedback. Typical approaches to predictive quantization and interpolation do not exploit the joint time-frequency correlation to enhance the precoder estimate at the transmitter. However, exploiting the correlations jointly not only reduces the feedback requirement, it also is able to enhance the precoder reconstruction accuracy. Further, the independence of the scalar parameters reduces the problem to interpolation of separate scalar valued function, as opposed to operations over tangent spaces in manifolds~\cite{Gupt1905:Predictive}. Simulations reveal that the proposed method offers matches the performance of manifold based approaches with a much lower implementation complexity and fewer number of bits.

The rest of this paper is organized as follows: Section~\ref{section2} described the system model and the precoder feedback and interpolation techniques. Section~\ref{section3} compares the BER and achievable rates for our proposed methods with that of past work. Finally, Section~\ref{section4} concludes with discussions on future work.

@@

In this paper, we propose a method for predictive quantization in the time domain over the subcarriers of a MIMO-OFDM system using scalar parameters. We also exploit the joint time-frequency correlation of these scalar parameters on the transmitter side. TODO how we solve it?

The rest of this paper is organized as follows: Section~\ref{section2}, we start by explaining the system model and the method for scalar quantization \& interpolation of the precoder matrix. In \ref{section3}, we perform various simulations for our approach and show the benefits from the improved BER rates and Achievable Rate.
%Since $H=U\Sigma V$, it is sufficient to quantize V*(precoder) and send this information back. In an OFDM channel, this data consume significant amount of bits and therefore, sending the complete data could consume too much bandwidth data. For e.g. an OFDM channel with 4 transmitters, 2 receiver antennas, and 64 subcarriers would require 64*(no. of bits required to send one precoder ~ 12-16 bits) ~ 896 bits per frame. This information could be significantly reduced if we provide better quantization and interpolation techniques/strategies. Also, the matrices which need to be fed back could be broken down into independent scalar parameters. These scalar quantities could be later quantized and interpolated.
% Broad summary of your work, come to your problem definition.

\section{System Model}
For a MIMO-OFDM system with $N_T$ transmitter and $N_R$ receiver
antennas, each subcarrier $i$ with time instant $t$ could be modelled as a flat-fading MIMO channel. This flat-fading MIMO channel can be modelled as
\label{section2}
\begin{equation}
\by_{i,t} = \bH_{i,t}\bV_{i,t} \bx_{i,t}+ \bf{\eta_{i,t}}
\end{equation}
where the channel matrix is
$\bH_{i,t} \in \mathbb{C}^{N_R \times N_T}$; the input data vector is
$\bx_{i,t} \in \mathbb{C}^{N_d}$; the output is
$\by_{i,t} \in \mathbb{C}^{N_R}$; the precoder matrix is
$\bV_{i,t} \in \mathbb{C}^{N_T \times N_R}$; and the complex additive white
Gaussian noise is $\eta$ whose distribution is
${\mathcal{CN}}(0,I_{N_R})$. We consider the case where $N_R <
N_T$ and the data vector size $N_d = N_R$.
The thin SVD of $\bH$ is given by
$\bH_{i,t} = \bU_{i,t} \Sigma_{i,t} \bV_{i,t}^{H}$, where
$\bU_{i,t} \in \mathbb{C}^{N_R \times N_R}$,
$\Sigma_{i,t} \in \mathbb{C}^{N_R \times N_R}$ and
$\bV_{i,t} \in \mathbb{C}^{N_T \times N_R}$. $\bU_{i,t}$ and
$\bV_{i,t}$ are matrices whose columns are orthonormal, and
$\Sigma_{i,t}$ contains the singular values
$\sigma_1 \geq \ldots \geq \sigma_{N_T} > 0$ of $\bH_{i,t}$. The
entries of $\bH_{i,t}$ are assumed to be distributed
i.i.d. $\tilde{\mathcal{N}}(0,1)$. To obtain a precoder at the
transmitter, we need to quantize and feed back information about
$\bV_{i,t}$ for each subcarrier.

% Givens rotation definition and how we are decomposing, Theta- pdf = 2lsin(theta)(2l-1).\cos(theta) 0 $\leq$ Theta $\leq$ pi/2.
% Householders' transformation and how we are decomposing. -pi$\leq$ phi $\leq$ pi. Uniformly distributed.
% Low-complexity quantization scheme based on a further parameterization of a vector in
% terms of a minimal number of independent scalar parameters. In particular, we consider its application to CSI feedback in slowly time-varying MIMO channels

\subsection{Scalar Parametrization}
\label{givens}
The degrees of freedom in the $\bV_{i,t}$ matrices is smaller than the
number of real entries in the matrix because of the geometrical
structure between the columns of the matrix~\cite{4114278}. The
orthonormal matrix $\bV_{i,t}$ with $N_T$ rows and $N_R$ columns can be
decomposed as TODO
\begin{equation}
\textbf{V} = \left[\prod_{k=1}^{N_{R}} \bD_{k} \left( \phi_{k,k},\ldots , \phi_{k,N_{R}} \right) \:  \prod_{l=1}^{N_{T}-k} \bG_{N_{T}-l,N_{T} -l+1} \big( \theta_{k,l}\big)  \right] \: \tilde{\bI}
\end{equation}
where $D_{k}$ is the diagonal matrix given as $D_{k}\big(\phi_{k,k},
\ldots, \phi_{k,N_T } \big)$ =  $\mbox{diag}\big( {\bf 1}_{k-1},
e^{j\phi_{k,k}},\ldots, e^{j\phi_{k,N_T }}  \big)$, ${\bf 1}_{k-1}$ is
represents  $k-1$ ones, and the $N_R \times N_T$ matrix $\tilde{\bI}$
= $\big[\bI_{N_R }, {\bf 0}_{N_T ,N_T -N_R }\big]^{T}$, and
\begin{equation}
G_{m-1,m}\big(\theta\big)  =
\begin{bmatrix}
I_{m-2} & & & \\
& \cos\theta & -\sin\theta & \\
& \sin\theta & \cos\theta & \\
& & & I_{t-m}
\end{bmatrix}
\end{equation}
Thus, for a $4 \times 2$ orthogonal matrix $\bV_{i,t}$, we have
\begin{align*}
  \bV_{i,t} & =
  \bD_{1}(\phi_{1,1},\ldots,\phi_{1,4})\bG_{3,4}(\theta_{1,1})
  \bG_{2,3}(\theta_{1,2}) \bG_{1,2}(\theta_{1,3})\times\\
& \bD_{2}(\phi_{2,2},\phi_{2,3},\phi_{2,4}) \bG_{3,4}(\theta_{2,1}) \bG_{2,3}(\theta_{2,2})\tilde{\bI}
\end{align*}
Here, $\phi_{k}$s are referred to as phases; their collection is given
by $\Phi$ and $\theta_{l}$s as rotation angles; their collection is
given by $\Theta$. $\phi_k \in (-\pi, + \pi]$ is uniformly distributed
between the two extremes~\cite{4114278}, while $\theta_l$ is
distributed as
\begin{equation}
2l(\sin\theta_l)^{2l-1}\cos\theta_l, \theta_l \in \left[0, \frac{\pi}{2}\right) \mbox{ , } l = 1,2,\ldots,N_T -1
\end{equation}
The $\phi_k$ and $\theta_l$ are statistically independent, making this
representation useful for tracking them separately across time, as
explained in the following subsection. The total number of parameters
obtained from the decomposition of a complex orthogonal matrix
$N_{T} \times N_{R} $ is $N_{R}(2N_{T} - N_{R})$~\cite{4114278}. The
number of $\phi_i$ parameters is $N_{R}(2N_{T} - N_{R}-1)/2$ while the
number of $\theta$ parameters is $N_{T}(2N_{T} - N_{R}+1)/2$.
\begin{figure}
\includegraphics[width=0.5\textwidth]{images/new-adpm.pdf}
\caption{\label{fig:adpm-fig}Subcarriers which need to be quantized}
\label{ber_overvie}
\vspace{-5pt}
\end{figure}
% To exploit the interpolation of time \& frequency jointly at the transmitter, feedback bits are sent in an alternating fashion as shown in the figure.

\subsection{Differential quantization and channel tracking}
\label{quantiz}
For channels that vary slowly with time, adaptive differential
quantization is an effective method for tracking parameters over time
using very few bits (often just one bit). For example, while tracking
a slowly varying discrete random process $x_n$ at time $n$, the one
bit can be used to determine the direction in which to move in order
to track the parameter. This can be denoted by
$\beta_{n} = \mbox{sign}(x_{n} - \hat{x}_{n-1})$, where $x_n$ is the
unquantized value and $\hat{x}_{n-1}$ is the unquantized (accurate)
value. Further, the step size for the move can be adapted based on the
speed of channel variation as well the speed for convergence. If
$\Delta_n$ is the step size, we define our adaptive quantizer as
\begin{equation}
\hat{x}_{n} = \hat{x}_{n-1} + \beta_{n}\Delta_{n} \mbox{ where }
\label{delta_eqn}
\Delta_{n} = \begin{cases}
    m \Delta_{n-1}, & \text{if $\beta_{n} = \beta_{n-1}$}\\
    \Delta_{n-1}/M , & \text{if $\beta_{n} \neq \beta_{n-1}$}.
  \end{cases}
\end{equation}
where $M$ and $m$ are the jumps used during the positive and negative
moves. We note that, in general, $m = M$, however, we have found that
using $M > m$ seems to yield better performance. We initialize
$\Delta_1$ as $\Delta_1 = |x_{2}-\hat{x}_1|$. We note that this is
performed separately for each of the quantized parameters.

@@

For the first and the second simulation time instances, since the
precoder is not available at the transmitter, we use 10 bits for
quantizing each parameter for effective representation of the
precoder. Therefore, the number of bits for each subcarrier =
$10\times12$ = 120. Quantization is performed as given in
Section~\ref{section2}. We note that this is not significant, since,
amortized over long durations, this number becomes small. Subsequent
quantization takes place in the time-domain using one bit for each
parameter. Since the bit budget is small, if the channel varies
sufficiently slowly, either the $theta_l$ or the $phi_i$ parameters
are fed back alternately, while retaining the unsent parameters as
being to their previous values, as shows in
Fig.~\ref{fig:adpm-fig}. Here, $\Theta$ is the the collection of all
$\theta_l$s and $\Phi$ is the collection of all $\phi_i$s. This brings
down the bit budget to an average of 5 bits(the length of $\Theta$)
and 7 bits (the length of $\Phi$) over alternate feedback instances,
thereby resulting in an average feedback budget of 5 bits.

To enable adaptive tracking of the channel parameters, we start with
an initialization of the step size for each parameter, as discussed in
Equation~\ref{delta_eqn}, with $m = 0.56, M = 1.4$ (the choice is
motivated by the results in~\ref{5671092}), and track parameters with
the temporal evolution of the channel. We have found that using
$M / m = 2.5$ yields optimum performance in terms of channel tracking.

Performance Measurements:
Since the Precoding matrices fundamentally lie on the Stiefel manifold, therefore, we are using the chordal distance parameter to measure the effectiveness of the quantization method used by us. Later we are comparing the BER rates with the ideal completely fed back subcarriers.

%talk about this in the simulations, that would be more relevant
Since $\Delta_{n} = M\Delta_{n-1} or \Delta_{n-1}/m $, we found empirically that $m=2.5*M$ i.e. decay

@@

\subsection{Performance Measurement}

\label{setting}

\noindent Since the Precoding matrices fundamentally lie on the Stiefel manifold, therefore, we are using the chordal distance parameter to measure the effectiveness of the quantization method used. Later we are comparing the BER rates with the completely fed back subcarriers. Or simulations and discussions part 2 criteria could be used. Done similar work as in \cite{4114278} but along with that we have joint interpolated in time and frequency at the transmitter using the fundamental scalar parameters obtained. To lower the number of bits we have used the method of interpolation and have achieved significant improvement over predictive quantization method in \cite{6891198} and also in \cite{Gupt1905:Predictive} which tried to use joint interpolation over the tangent space in the Stiefel Manifold. In fact, we use joint interpolation of the scalar parameters which are easy to use and give better quantization than any of the above methods.

>>>>

In this paper, we propose a method for exploiting the joint time-frequency correlation of the precoder's scalar parameters on the transmitter using CSI feedback. Typical approaches to predictive quantization and interpolation do not exploit the joint time-frequency correlation to enhance the precoder estimate at the transmitter. However, exploiting the correlations jointly not only reduces the feedback requirement, but also enhances the precoder reconstruction accuracy significantly. Further, the independence of the scalar parameters reduces the problem to interpolation of separate scalar valued function, as opposed to operations over tangent spaces in manifolds~\cite{Gupt1905:Predictive}. Simulations reveal that the proposed method offers matches the performance of manifold based approaches with a much lower implementation complexity and fewer number of bits.

The rest of this paper is organized as follows: Section~\ref{section2} describes the system model and the precoder feedback and interpolation techniques. Section~\ref{section3} compares the BER and achievable rates for our proposed methods with other approaches. Finally, Section~\ref{section4} concludes with discussions on future work.

@@

% The rest of this paper is organized as follows: Section~\ref{section2}, we start by explaining the system model and the method for scalar quantization \& interpolation of the precoder matrix. In \ref{section3}, we perform various simulations for our approach and show the benefits from the improved BER rates and Achievable Rate.
%Since $H=U\Sigma V$, it is sufficient to quantize V*(precoder) and send this information back. In an OFDM channel, this data consume significant amount of bits and therefore, sending the complete data could consume too much bandwidth data. For e.g. an OFDM channel with 4 transmitters, 2 receiver antennas, and 64 subcarriers would require 64*(no. of bits required to send one precoder ~ 12-16 bits) ~ 896 bits per frame. This information could be significantly reduced if we provide better quantization and interpolation techniques/strategies. Also, the matrices which need to be fed back could be broken down into independent scalar parameters. These scalar quantities could be later quantized and interpolated.
% Broad summary of your work, come to your problem definition.

\section{System Model}
For a MIMO-OFDM system with $N_T$ transmitter and $N_R$ receiver
antennas, each subcarrier $i$ with time instant $t$ could be modelled as a flat-fading MIMO channel. This flat-fading MIMO channel can be modelled as
\label{section2}
\begin{equation}
\by_{i,t} = \bH_{i,t}\tilde{\bV}_{i,t} \bx_{i,t}+ \bf{\eta_{i,t}}
\end{equation}
where the channel matrix is
$\bH_{i,t} \in \mathbb{C}^{N_R \times N_T}$; the input data vector is
$\bx_{i,t} \in \mathbb{C}^{N_d}$; the output is
$\by_{i,t} \in \mathbb{C}^{N_R}$; the unitary precoder matrix is
$\tilde{\bV}_{i,t} \in \mathbb{C}^{N_T \times N_R}$; and the complex
additive white Gaussian noise is $\eta$ whose distribution is
${\mathcal{CN}}(0,I_{N_R})$. We consider the case where $N_R < N_T$
and the data vector size $N_d = N_R$.  The ``thin'' SVD (that keeps
only those right and left singular vectors that correspond to non-zero
singular values) of $\bH$ is given by
$\bH_{i,t} = \bU_{i,t} \Sigma_{i,t} \bV_{i,t}^{H}$, where
$\bU_{i,t} \in \mathbb{C}^{N_R \times N_R}$,
$\Sigma_{i,t} \in \mathbb{C}^{N_R \times N_R}$ and
$\bV_{i,t} \in \mathbb{C}^{N_T \times N_R}$. $\bU_{i,t}$ and
$\bV_{i,t}$ are matrices whose columns are orthonormal, and
$\Sigma_{i,t}$ contains the singular values
$\sigma_1 \geq \ldots \geq \sigma_{N_T} > 0$ of $\bH_{i,t}$. The
entries of $\bH_{i,t}$ are assumed to be distributed
i.i.d. $\tilde{\mathcal{N}}(0,1)$. To obtain a precoder at the
transmitter, we need to quantize and feed back information about
$\bV_{i,t}$ for each subcarrier.

\subsection{Scalar Parametrization}
\label{givens}
The degrees of freedom in the $\tilde{\bV}_{i,t}$ matrices is smaller
than the number of real entries in the matrix because of the
geometrical structure between the columns of the
matrix~\cite{4114278}. The orthonormal matrix $\tilde{\bV}_{i,t}$ with
$N_T$ rows and $N_R$ columns can be decomposed as follows~\cite{4114278} (we suppress
the $(i, t)$ subscripts in the decomposition matrices)
\begin{equation}
\tilde{\bV}_{i,t} = \left[\prod_{k=1}^{N_{R}} \bD_{k} \left( \phi_{k,k},\ldots , \phi_{k,N_{R}} \right) \:  \prod_{l=1}^{N_{T}-k} \bG_{N_{T}-l,N_{T} -l+1} \big( \theta_{k,l}\big)  \right] \: \tilde{\bI}
\end{equation}
where $\bD_{k}$ is a diagonal matrix defined as
$\bD_{k}\big(\phi_{k,k}, \ldots, \phi_{k,N_T } \big)$ =
$\mbox{diag}\big( {\bf 1}_{k-1}, e^{j\phi_{k,k}},\ldots,
e^{j\phi_{k,N_T }} \big)$, ${\bf 1}_{k-1}$ is represents $k-1$ ones,
and the $N_R \times N_T$ matrix $\tilde{\bI}$ =
$\big[\bI_{N_R }, {\bf 0}_{N_T ,N_T -N_R}\big]^{T}$. Here,
${\bf 0}_{N_T ,N_T -N_R }$ is an $N_T\times (N_T - N_R)$ matrix all of
whose elements are zero. Finally, we also have
\begin{equation}
\bG_{m-1,m}\big(\theta\big)  =
\begin{bmatrix}
I_{m-2} & & & \\
& \cos\theta & -\sin\theta & \\
& \sin\theta & \cos\theta & \\
& & & I_{t-m}
\end{bmatrix}
\end{equation}
Thus, for a $4 \times 2$ orthogonal matrix $\bV_{i,t}$, we have
\begin{align*}
  \bV_{i,t} & =
  \bD_{1}(\phi_{1,1},\ldots,\phi_{1,4})\bG_{3,4}(\theta_{1,1})
  \bG_{2,3}(\theta_{1,2}) \bG_{1,2}(\theta_{1,3})\times\\
& \bD_{2}(\phi_{2,2},\phi_{2,3},\phi_{2,4}) \bG_{3,4}(\theta_{2,1}) \bG_{2,3}(\theta_{2,2})\tilde{\bI}
\end{align*}
Here, $\phi_{k}$s are referred to as phases, and their collection for
the $i$-th subcarrier at time instant $t$ is captured in a vector
denoted by $\Phi_{i,t}$. $\theta_{l}$s as called rotation angles, and
their collection is captured in a vector referred to as
$\Theta_{i,t}$. We note that $\Phi_{i,t}$ and $\Theta_{i,t}$ represent
the phase and rotation angles for the $i$-th subcarrier at time
instant $t$, as represented in Fig.~\ref{fig:adpm-fig}. It is known
that $\phi_k \in (-\pi, + \pi]$ is uniformly distributed between the
two extremes~\cite{4114278}, while $\theta_l$ is distributed as
\begin{equation}
2l(\sin\theta_l)^{2l-1}\cos\theta_l, \theta_l \in \left[0, \frac{\pi}{2}\right) \mbox{ , } l = 1,2,\ldots,N_T -1
\end{equation}
The $\phi_k$ and $\theta_l$ are statistically independent, making this
representation useful for tracking them separately across time, as
explained in the following subsection. The total number of parameters
obtained from the decomposition of a complex orthogonal matrix
$N_{T} \times N_{R} $ is $N_{R}(2N_{T} - N_{R})$. The
number of $\phi_k$ parameters is $N_{R}(2N_{T} - N_{R}-1)/2$ while the
number of $\theta$ parameters is $N_{R}(2N_{T} - N_{R}+1)/2$. Thus, $\Phi_{i,t}$ is a vector of length $N_{R}(2N_{T} - N_{R}-1)/2$, while $\Theta_{i,t}$ has length $N_{R}(2N_{T} - N_{R}+1)/2$.
\begin{figure}
\includegraphics[width=0.5\textwidth]{images/new-adpm.pdf}
\caption{\label{fig:adpm-fig}Subcarriers which need to be quantized}
\end{figure}
% To exploit the interpolation of time \& frequency jointly at the transmitter, feedback bits are sent in an alternating fashion as shown in the figure.

\subsection{Differential quantization and channel tracking}
\label{quantiz}
For channels that vary slowly with time, adaptive differential
quantization is an effective method for tracking parameters over time
using very few bits (often just one bit per parameter). For example,
while tracking a slowly varying discrete random process $x_n$ at time
$n$, the one bit can be used to determine the direction in which to
move in order to track the parameter. This can be denoted by
$\beta_{n} = \mbox{sign}(x_{n} - \hat{x}_{n-1})$, where $x_n$ is the
unquantized value and $\hat{x}_{n-1}$ is the unquantized (accurate)
value. Further, the step size for the move can be adapted based on the
speed of channel variation as well the speed for convergence. If
$\Delta_n$ is the step size, we define our adaptive quantizer as
\begin{equation}
\hat{x}_{n} = \hat{x}_{n-1} + \beta_{n}\Delta_{n} \mbox{ where }
\label{delta_eqn}
\Delta_{n} = \begin{cases}
    M \Delta_{n-1}, & \text{if $\beta_{n} = \beta_{n-1}$}\\
    \Delta_{n-1}/M , & \text{if $\beta_{n} \neq \beta_{n-1}$}.
  \end{cases}
\end{equation}
where $M$ is the increment used during the positive and negative
moves. We initialize $\Delta_1$ as $\Delta_1 = |x_{2}-\hat{x}_1|$. We
note that this is performed separately for each of the quantized
parameters.

@@

For the first and the second simulation time instances, since the
precoder is not available at the transmitter, we use 10 bits for
quantizing each parameter for effective representation of the
precoder. Therefore, the number of bits for each subcarrier =
$10\times12$ = 120. Quantization is performed as given in
Section~\ref{section2}. We note that this is not significant, since,
amortized over long durations, this number becomes small. Subsequent
quantization takes place in the time-domain using one bit for each
parameter. Since the bit budget is small, if the channel varies
sufficiently slowly, either the $\theta_l$ or the $\phi_i$ parameters
are fed back alternately, while retaining the unsent parameters as
being to their previous values, as shows in
Fig.~\ref{fig:adpm-fig}. Here, $\Theta$ is the the collection of all
$\theta_l$s and $\Phi$ is the collection of all $\phi_i$s. This brings
down the bit budget to an average of 5 bits(the length of $\Theta$)
and 7 bits (the length of $\Phi$) over alternate feedback instances,
thereby resulting in an average feedback budget of 5 bits.

To enable adaptive tracking of the channel parameters, we start with
an initialization of the step size for each parameter, as discussed in
Equation~\ref{delta_eqn}, with $M = 1.4$ (the choice is motivated by
the results in~\cite{5671092}), and track parameters with the temporal
evolution of the channel, and we find that this choice effectively
tracks the channel parameters.

% Performance Measurements:
% Since the Precoding matrices fundamentally lie on the Stiefel manifold, therefore, we are using the chordal distance parameter to measure the effectiveness of the quantization method used by us. Later we are comparing the BER rates with the ideal completely fed back subcarriers.

% Using the idea of joint interpolation given in the other paper.

% %talk about this in the simulations, that would be more relevant
% Since $\Delta_{n} = M\Delta_{n-1} or \Delta_{n-1}/m $, we found empirically that $m=2.5*M$ i.e. decay

@@

Since the Precoding matrices fundamentally lie on the Stiefel
manifold, therefore, we are using the chordal distance parameter to
measure the effectiveness of the quantization method used. Later we
are comparing the BER rates with the completely fed back
subcarriers. Or simulations and discussions part 2 criteria could be
used. Done similar work as in \cite{4114278} but along with that we
have joint interpolated in time and frequency at the transmitter using
the fundamental scalar parameters obtained. To lower the number of
bits we have used the method of interpolation and have achieved
significant improvement over predictive quantization method in
\cite{6891198} and also in \cite{Gupt1905:Predictive} which tried to
use joint interpolation over the tangent space in the Stiefel
Manifold. In fact, we use joint interpolation of the scalar parameters
which are easy to use and give better quantization than any of the
above methods.
